A large leading tail of atmospheric escape in the HAT-P-67 b system

### 2023-03-09 paper/paper1/ms.tex
Introduction threads the story now

Uncertainty in system ages, evaporation timescales, and dominating physical mechanisms, degrade our ability to foretell if any given planet will exhibit ongoing signatures of atmospheric escape.  Episodic stellar wind gusts and other forms of cosmic variance could also subdue atmospheric escape, even where we expect it most.  The 48 published non-detections of atmospheric escape \citep{2022arXiv221116243D} must encode these natural whims in a way that we have not yet disentangled. Nevertheless, we can boost our chances of witnessing active and significant atmospheric escape by targeting sources with favorable traits.  These traits may include close-in, low surface gravity exoplanets within a strategically chosen radius range.  The 1.5-2.0 Earth radius range could be expected to harbor atmospheres that have not yet shed their gaseous envelopes, still in the act of eroding to their final rocky state.  Alternatively, inflated Hot Jupiters could be so dramatically ablated as to overcome the somewhat stronger potential well and facilitating hydrodynamic flows.  Significant gains in understanding can therefore be expected from observational studies sampling the extremes of the exoplanet distributions.

Inflated hot Saturns stand out as an especially extreme category that should exhibit mass loss.  Their low gravitational potentials should let go of their atmospheres more easily than their more steadfast Hot Jupiter counterparts.  Their lower gravity yields larger atmospheric scale heights making them easier to detect in transmission spectroscopy.  And their large transit depths and short periods should make them easy to detect in large numbers, like Hot Jupiters.

@@

Alternatively---and most consequentially for atmospheric escape---another explanation may prevail.  Inflated hot Saturns may effectively migrate to close-in orbital separations, but once they arrive the intense irradation overheats the planet, leading to runaway inflation and ultimately complete disintegration.  The atmosphere overflows its Roche lobe, the potential well softens, the Roche lobe shrinks, and a positive feedback loop ensues. The mass loss increases expoentially.  In this scenario, the inflationary half-life may be so short as to lead to a low probability of observing members in the class, and hence the apparent lack of inflated sub-Saturns.

\citet{2011ApJ...738....1B} predicted runaway inflation of hot Saturns as a consequence of the Ohmic dissipation mechanism.  Ohmic dissipation

\citet{2018AJ....155..214T} stated these two explanations for the lack of sub-Saturns, without a way to distinguish which one was at play.

This inflationary scenario would predict a profound mass loss rate for inflated hot Saturns.  If the planet is inflated, it is already in the runaway mass loss phase.

Here we present a study of the exosphere of \object{HAT-P-67 b}, an inflated hot Saturn orbiting an F type sub-giant\footnote{The system appears to consist of a binary with an M dwarf companion HAT-P-67B separated on-sky by $9\farcs0$ \citep{2019MNRAS.490.5088M}.  The IAU naming convention would demand HAT-P-67Ab.  We simply drop the A designation for notational simplicity, since the wide companion will not factor into our analysis.} at a distance of 0.06 AU \citep{2017AJ....153..211Z}.  The strong insolation combined with \object{HAT-P-67 b}'s low surface gravity makes it an excellent candidate for strong atmospheric mass loss.  It orbits an evolved star, opening the prospect for the planet's insolation to have increased over time as the star grew.  These and other reasons make \object{HAT-P-67 b} a predestined laboratory for measuring the properties of exospheres at the extreme.

Candidates for detection should usually be observable in one of three atomic lines.  Ly~$\alpha$, \ion{He}{1} 10833 \AA, and H~$\alpha$ have emerged as the most amenable to detection \citep{2018ApJ...855L..11O,2023MNRAS.518.4357O}.  Ly $\alpha$ can exhibit strong atmospheric escape signals, but instellar medium \ion{H}{1} absorption censors its low-velocity linecore.  \ion{He}{1} 10833 \AA~ offers significant advantages over Ly~$\alpha$, and has resulted in at least 14 systems with detections to date, including \object[GJ 3470]{GJ3470} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  These detections exhibit a range of signal strengths, as high as 10$\%$ transit depths.  Measurements at high spectral resolution have been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}

@@

\subsection{Gaia DR3}
HAT-P-67A (\emph{Gaia DR3 1358614983131339392}) has a parallax of $2.69\pm0.01$ mas in \emph{Gaia} DR3, placing it at about 372$\pm$1.4 pc, slightly farther than previously estimated by \citet{2017AJ....153..211Z} using Gaia DR1. Its coarse Gaia-based estimates for RV and Doppler broadening are consistent with the more precise values published in \citet{2017AJ....153..211Z}.  Its wide companion HAT-P-67B (\emph{Gaia DR3 1358614983131339904}) \citep{2019MNRAS.490.5088M} has statistically identical parallax ($2.58\pm0.05$) and proper motions, confirming its interpretation as co-moving, with projected separation of about 3400 AU.

>>>>

Uncertainty in system ages, evaporation timescales, and dominating physical mechanisms, degrade our ability to foretell if any given planet will exhibit ongoing signatures of atmospheric escape.  Episodic stellar wind gusts and other forms of cosmic variance could also subdue atmospheric escape, even where we expect it most.  The 48 published non-detections of atmospheric escape \citep{2022arXiv221116243D} must encode these natural whims in a way that we have not yet disentangled. Nevertheless, we can boost our chances of witnessing active and significant atmospheric escape by targeting sources with favorable traits.  These traits may include close-in, low surface gravity exoplanets within a strategically chosen radius range.  For example, the 1.5-2.0 Earth radius range could be expected to harbor atmospheres that have not yet shed their gaseous envelopes, still in the act of eroding to their final rocky state.

Inflated hot Saturns stand out as an especially extreme category that should exhibit mass loss.  Their low gravitational potentials should let go of their atmospheres more readily than their steadfast Hot Jupiter counterparts.  Their lower gravity yields larger atmospheric scale heights making them easier to detect in transmission spectroscopy.  And their large transit depths and short periods should make them easy to detect in large numbers, like Hot Jupiters.

@@

Alternatively---and most consequentially for atmospheric escape---another explanation may prevail.  Inflated hot Saturns may effectively migrate to close-in orbital separations, but once they arrive, the intense irradation overheats the planet.  This heating leads to runaway inflation and ultimately complete disintegration.  In this scenario, the inflationary half-life may be so short as to lead to a low probability of observing members in the class, and hence the apparent lack of inflated sub-Saturns.

\citet{2011ApJ...738....1B} predicted runaway inflation of hot Saturns as a consequence of the Ohmic dissipation mechanism.  Here, a lightly ionized atmosphere experiences drag in a planetary magnetic field, weakly coupling the stellar incident energy into the planetary interior.  \citet{2018AJ....155..214T} proposed these two explanations for the lack of sub-Saturns, while statistically favoring Ohmic dissipation as the mechanism responsible for inflating Hot Jupiters.  Ohmic dissipation does not preclude the existence and role of other heating mechanisms---such as photoevaporation and tidal heating---but it does stand out has having made these predictions before data were widely available to test these predictions.

The runaway inflation scenario could yield a profound mass loss rate for inflated hot Saturns, possibly much larger than we have seen in previous systems.  The ability to detect that mass loss hinges on their observability in spectral tracers.  Ly~$\alpha$, \ion{He}{1} 10833 \AA, and H~$\alpha$ have emerged as the most amenable to detection \citep{2018ApJ...855L..11O,2023MNRAS.518.4357O}, but each of these has its own limitations.  Ly $\alpha$ can suffer from instellar medium \ion{H}{1} absorption censoring its low-velocity linecore, for example.  Here we focus on \ion{He}{1} 10833 \AA~, which offers some advantages over Ly~$\alpha$.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{H}{1} has resulted in at least 14 systems with detections, including \object[GJ 3470]{GJ3470} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  To date, no inflated hot Saturns appear to have been examined, neither in Helium nor other tracers \emph{c.f.} Figure \censorbox{figure X}.

Here we present a multi-year observational campaign searching for atmospheric escape in \object{HAT-P-67 b}, an inflated hot Saturn orbiting an F type sub-giant at an orbital separation of 0.06 AU \citep{2017AJ....153..211Z}.  The strong insolation combined with \object{HAT-P-67 b}'s low surface gravity makes it an excellent candidate for strong atmospheric mass loss.  It orbits an evolved star, opening the prospect for ``re-inflation'' as the planet's insolation increased over time as the star expanded.  Its status as a rare inflated Hot Saturn make \object{HAT-P-67 b} a promising laboratory for testing the inflationary predictions of the Ohmic dissipation mechanism.

@@

\subsection{Gaia DR3}
The stellar system consists of a binary with an M dwarf companion HAT-P-67B separated on-sky by $9\farcs0$ \citep{2019MNRAS.490.5088M}.  The IAU naming convention would demand HAT-P-67A\emph{b} to refer to the planet.  HAT-P-67A (\emph{Gaia DR3 1358614983131339392}) has a parallax of $2.69\pm0.01$ mas in \emph{Gaia} DR3, placing it at about 372$\pm$1.4 pc, slightly farther than previously estimated by \citet{2017AJ....153..211Z} using Gaia DR1. Its coarse Gaia-based estimates for RV and Doppler broadening are consistent with the more precise values published in \citet{2017AJ....153..211Z}.  Its wide companion HAT-P-67B (\emph{Gaia DR3 1358614983131339904}) \citep{2019MNRAS.490.5088M} has statistically identical parallax ($2.58\pm0.05$) and proper motions, confirming its interpretation as co-moving, with projected separation of about 3400 AU.  Hereafter, we simply drop the A designation for notational simplicity, since the wide companion will not factor into our analysis.
